Argennon: A Scalable Smart Contract Platform

### 2022-07-01 tex/sections/avm.tex
Fix some readability issues
Signed-off-by: aybehrouz <behrouz_ayati@yahoo.com>

The Argennon\footnote{The classical pronunciation should be used:\textipa{/Ar"gen.non/}} Smart Contract Execution
Environment (AscEE) is an abstract execution environment for executing Argennon smarts contracts (a.k.a\ Argennon
applications). An Argennon application essentially is an HTTP server whose state is kept in the Argennon blockchain and
its logic is described using an Argennon Standard Representation (ASR).

An Argennon Standard Representation (ASR) is a programming language for describing argennon applications, optimized
for the architecture and requirements of the Argennon platform.
Argennon supports two standard representations. One representation is a high level text based language which needs
costly compilation before being executed on a hardware machine. The other, is a low level binary representation which
usually can be executed, with minimal pre-processing, by a JIT compiler or an emulator. The high level
language is intended for preserving the high level information of applications logic to facilitate
platform specific compiler optimization at host nodes. The low level language, on the oder hand, is designed for
efficient direct execution of applications that are not frequently used.

@@

The Argennon Smart Contract Execution Environment can be seen as a machine for executing Argennon applications to
fulfill HTTP requests, produce their HTTP responses and update related heap chunks. The execution of requests can be
considered sequential\footnote{Actually requests are executed in parallel but by performing data dependency analysis the
result is guaranteed to be identical
with sequential execution of requests.} and each request has a separate \emph{execution session}. An execution session
is a separate session of executing smart contract's code in order to fulfill an \emph{external} HTTP request.
External requests are requests that are not made by other Argennon applications.

@@

The AscEE is designed based on \emph{optional decoupling principle}. When an application makes a request to another
application, optionally it can choose to be decoupled from the called application. That would mean the called
application could not affect its caller's state by reentrancy, or could not fail the session by using excessive
resources or performing illegal operations.

@@

An Argennon identifier may be simple or compound. A simple identifier is generated using a single trie, while a
compound identifier is generated by concatenating prefix codes generated by two or more tries:

@@

All Argennon prefix trees have an equal branching factor \(\beta\). Therefore, we can represent an Argennon
prefix tree as a sequence of fractional numbers\footnote{It's possible to have \(a_i=0\).
For example, \(A^{(4)}=(0.2000)_{10}\) is correct.} in base \(\beta\):
\[
    (A^{(1)},A^{(2)},A^{(3)},\dots)\ ,
\]
where \(A^{(i)}=(0.a_{1}a_{2}\dots a_{i})_\beta\), and we have \(A^{(i)} \leq A^{(i+1)}\).\footnote{A typical choice for
    \(\beta\) is \(2^8\).}

@@

The persistent data of an Argennon application is stored in heap chunks. An application may have several heap chunks
with different sizes, and can remove or
resize its chunks or allocate new chunks. Every chunk belongs to exactly one application. Only the owner application can
modify a chunk but there is no restrictions for reading a chunk\footnote{The reason behind this type of access
control design is the fact that smart contract
code is usually immutable. That means if a smart contract does not implement a
getter mechanism for some parts of its internal data, this functionality can never
be added later, and despite the internal data is publicly available, there will be no
way for other smart contracts to use this data on-chain.}.

@@

A heap chunk can be considered as a continuous
array of bytes. Every chunk has a size: \texttt{chunkSize} and a size upper bound: \texttt{sizeUpperBound}. The value of
\texttt{chunkSize} can be determined uniquely at
the start of
every execution session, and it may be updated during the session like a normal memory location. On the other hand,
the value of \texttt{sizeUpperBound} is constant for every block of the blockchain and is proposed by the block
proposer. \texttt{sizeUpperBound} indicates the upper bound of \texttt{chunkSize} value for a chunk in a block.
A validator can use this value to safely allocate memory for a chunk.

@@

Memory locations inside a chunk can only be accessed through access blocks. An access block is defined on a chunk
and has an offset and a size and determines accessible memory locations inside the chunk. Multiple access blocks can
be defined on a single chunk, but they must be non overlapping. Access blocks are byte addressable and can have
different access types:

\begin{itemize}
    \item \texttt{read\_only}: only allows read and check operations.
    \item \texttt{writable}: allows reading and writing.
    \item \texttt{check\_only}: only allows check operations. These operations query the persistence
    status of a memory location.
    \item \texttt{additive}: only allows addition-like operations without overflow checking. Note that the content of
    these access blocks cannot be read.
\end{itemize}

Chunks and access blocks have different purposes. Chunks are intended for simplifying proof checking of the data
stored in the Argennon cloud\footnote{Note that
the Argennon cloud uses cryptographic accumulators for preserving data integrity.} and access blocks are
required for better parallelization of the request execution. An application should put the data it predicts is
needed for validating a block in the same chunk and the data it predicts is needed in a single execution
session in the same access block.

@@

The AscEE uses \emph{Authenticated Message Passing} for authorizing operations. In this method, every execution
session has a set of authenticated messages, and applications \textbf{explicitly} pass those messages for authorizing
operations.
Authentication of messages can be efficiently done in parallel, using any type of cryptographic signature scheme in
the preprocessing phase of request execution. For example, we could use BLS aggregate signature to authenticate all
messages of a block in bulk.

Moreover, applications use built-in functions of AscEE to generate authenticated messages in run-time. This enables
an application to authorize operations for another application even if that application is not directly called by
itself.

@@

Resource usage can be measured per session or per application call. Obviously per session measurement is
easier and more efficient. However, when we measure resources per session if
a session violates its resource caps, determining the point of failure may require precise and error-free resource
measurement. For example, assume that a session containing an application call
violates a 2 milliseconds execution time cap. If in the caller's code, the call happens exactly after 2 milliseconds,
a small fluctuation in the execution time measurement can
change the point of failure between the called and the caller application. Note that for implementing optional
decoupling principle in Argennon, we need to determine the
exact application call which has failed in a call chain, and this fluctuation introduces a nondeterministic
behaviour which could make block validation impossible. That's why the AscEE sometimes uses per application call
resource management. If we define the resource caps per application call and
perform our measurements for each application call separately, by using caps that are larger than the measurement
error, errors in the measurements would not change the point of failure in the call chain.

The AscEE has two type of execution sessions: \emph{optimistic} and \emph{monitored}. Resource usage of an optimistic
session is always measured per session and default pre-defined resource caps are used. On the other hand, for a
monitored session, resources that can not be measured precisely are measured per application call and their caps are
determined per application call by the external HTTP request (i.e.\ transaction). The block proposer decides the
type of the execution session for each transaction.

In most consensus protocols, we can assume that the block proposer has enough incentive to filter out transactions
that spam run-time resources. Here by a run-time resource, we refer to a resource that at run-time, a limited amount
of it is available, but its surplus can not be stored for later use. Execution time and local memory are
examples of a run-time resource but permanent storage is not a run-time resource.

If a proposed block contains many transactions which need a lot of run-time resources, validators would not
be able to validate all transactions in a timely manner. Consequently, they may decide to reject the block or if they
spend enough resources, the confirmation of that block could take more than usual time. Longer block time is not
favoured by block proposers, because it means less throughput of the system which usually means less overall rewards
for them. In the Argennon protocol the management of run-time resources, is left to the block proposer.

Different computational resources are measured and monitored during an AscEE session:
\begin{itemize}
    \item \textbf{execution time}:
    is the amount of cpu time that is required for executing a session or an application call. The execution time is
    measured in \emph{AscEE clocks}. One AscEE clock is defined as 1/1000 of the amount
    of \textbf{cpu time} needed for executing a predefined standard application which is used for benchmarking a host's
    performance. In other words, by definition the AscEE imaginary standard machine completes the standard benchmark
    in 1000 clocks.

@@

In Argennon, execution time and local memory are considered \emph{nondeterministic} resources. A nondeterministic
resource is a resource that can not be measured precisely and its measurement always contains a random error.
Optimistic sessions are not allowed to fail because of violating nondeterministic resource limits. As a result, the
block proposer must always choose a monitored session for a transaction that is included in his block and violates a
nondeterministic resource cap.

When a transaction fails due to violation of a limit for a nondeterministic resource, the proposer is
required to exactly specify the application call in the session that violates that limit. When validators are
executing a monitored session, for each application call, they enforce considerably larger limits for
nondeterministic resources, only when the proposer has declared that an application call violates a limit, the
validators will enforce the actual value of the limit. This simple mechanism ensures that, with a very high
probability, validators agree with the proposer.

@@

An application can invoke methods of the ArgC Standard Library (ASL) in its own context. Methods of the ArgC standard
library are stored as a special application in the Argennon blockchain.
a part of the root smart contract.
In Argennon, some applications (smart contracts) are updatable. The ArgC Standard Library is an updatable smart
contract which can be updated by the Argennon governance
system. This means that bugs or security vulnerabilities in the ArgC Standard Library could be quickly patched and
applications could benefit from bugfixes and improvements of the ArgC Standard Library even if they are
non-updatable. Many important and useful functionalities,
such as fungible and non-fungible assets, access control mechanisms,
and general purpose DAOs are implemented in the ArgC Standard Library.

>>>>

The Argennon\footnote{The classical pronunciation should be used:\textipa{/Ar"gen.non/}} Smart Contract Execution
Environment (AscEE) is an abstract execution environment for executing Argennon smarts contracts (a.k.a\ Argennon
applications) in a safe and secure environment. An Argennon application essentially is an HTTP server whose state is
kept in the Argennon blockchain and its logic is described using an Argennon Standard Representation (ASR).

An Argennon Standard Representation (ASR) is a programming language for describing argennon applications, optimized
for the architecture and properties of the Argennon platform.
Argennon supports two standard representations: One is a high level text based representation which needs
costly compilation before being executed on a hardware machine. The other, is a low level binary representation which
usually can be executed, with minimal pre-processing, by a JIT compiler or an emulator. The high level
language is intended for preserving the high level information of applications logic to facilitate
platform specific compiler optimization at host nodes. The low level language, on the oder hand, is designed for
efficient direct execution of applications that are not frequently used and do not benefit from compiler optimizations.

@@

The Argennon Smart Contract Execution Environment can be seen as a machine for executing Argennon applications to
fulfill HTTP requests, produce their HTTP responses and update related heap chunks. The execution of requests can be
considered sequential\footnote{Actually requests are executed in parallel but by performing data dependency analysis the
result is guaranteed to be identical with sequential execution of requests.} and each request has a separate
\emph{execution session}. Therefore, an execution
session is a separate session of executing smart contract's code in order to fulfill an \emph{external} HTTP request.
External requests are requests that are not made by other Argennon applications.

@@

The AscEE is designed based on \emph{optional decoupling principle}. When an application makes a request to another
application, optionally it can choose to be decoupled from the called application. That would mean the called
application could not affect its caller's state by reentrancy, or could not abort the session by using excessive
resources or performing illegal operations.

@@

An Argennon identifier may be simple or compound. A simple identifier is generated using a single tree, while a
compound identifier is generated by concatenating prefix codes generated by two or more trees:

@@

All Argennon prefix trees have an equal branching factor \(\beta\)\footnote{A typical choice for \(\beta\) is \(2^8\)
    .}, and
we can represent an Argennon
prefix tree as a sequence of fractional numbers\footnote{It's possible to have \(a_i=0\).
For example, \(A^{(4)}=(0.2000)_{10}\) is correct.} in base \(\beta\):
\[
    (A^{(1)},A^{(2)},A^{(3)},\dots)\ ,
\]
where \(A^{(i)}=(0.a_{1}a_{2}\dots a_{i})_\beta\), and we have \(A^{(i)} \leq A^{(i+1)}\).

@@

The persistent data of an Argennon application is stored in continuous finite
arrays of bytes, called heap chunks. An application may have several heap chunks
with different sizes, and can remove or
resize its chunks or allocate new chunks. Every chunk belongs to exactly one application. Only the owner application can
modify a chunk but there is no restrictions for reading a chunk\footnote{The reason behind this type of access
control design is the fact that smart contract
code is usually immutable. That means if a smart contract does not implement a
getter mechanism for some parts of its internal data, this functionality can never
be added later, and despite the internal data is publicly available, there will be no
way for other smart contracts to use this data on-chain.}.

@@

Every heap chunk has a size: \texttt{chunkSize} and a size upper bound: \texttt{sizeUpperBound}. The value of
\texttt{chunkSize} can be determined uniquely at
the start of
every execution session, and it may be updated by the owner application during the session like a normal memory
location. On the other hand,
the value of \texttt{sizeUpperBound} is constant for every block of the Argennon blockchain and is proposed by the block
proposer. This value indicates the upper bound of \texttt{chunkSize} for a specific chunk during the execution of all
requests of a block, and it can be used by validators for memory allocation of chunks during block validation.

@@

Memory locations inside a chunk can only be accessed through access blocks. An access block is an access interval
defined on a chunk. It has an offset and a size and determines the accessible memory locations inside the chunk.
Multiple access blocks can be defined on a single chunk, but they must be non overlapping.
Access blocks are byte addressable and can have different access types:

\begin{itemize}
    \item \texttt{check\_only}: only allows check operations. These operations query the persistence
    status of a memory location.
    \item \texttt{read\_only}: only allows read and check operations.
    \item \texttt{writable}: allows reading and writing.
    \item \texttt{additive}: only allows additive operations. These operations are different types of mathematical
    addition without any overflow checking. Note that the content of these access blocks cannot be read.
\end{itemize}

Chunks and access blocks have different purposes. Chunks are intended for simplifying proof checking of the data
stored in the Argennon cloud\footnote{Note that
the Argennon cloud uses cryptographic accumulators for preserving data integrity.} and access blocks are
required for better parallelization of the request execution. An application should put the data needed for
validating a block in the same chunk and the data needed in a single execution session in the same access block.

@@

The AscEE uses \emph{Authenticated Message Passing} for authorizing operations. In this method, every execution
session has a set of authenticated messages, and those messages are \textbf{explicitly} passed in requests to
applications for authorizing operations. These messages act exactly like digital signatures and applications can
ensure that they are issued by specific accounts. The only difference is that the process of message authentication
is performed by the AscEE internally and is not performed by the application. For accounts, the AscEE uses digital
signatures to authenticate messages. Signatures for each request can be validated in parallel during the
preprocessing phase of the request execution, and any type of
cryptographic signature scheme can be used. For example, we could use BLS aggregate signature to authenticate all
messages of a block of the blockchain in bulk.

Moreover, applications can use built-in functions of the AscEE to generate authenticated messages in run-time.
This enables an application to authorize operations for another application even if it is not calling that
application directly.

@@

In most consensus protocols, we can assume that the block proposer has enough incentive to filter out transactions
that spam run-time resources. Here by a run-time resource, we mean a resource that at run-time, a limited amount
of it is available, but its surplus can not be stored for later use. Execution time and local memory are
examples of such a resource but permanent storage is not.

If a proposed block contains many transactions which need a lot of run-time resources, validators would not
be able to validate all transactions in a timely manner. Consequently, they may decide to reject the block or if they
spend enough resources, the confirmation of that block could take more than usual. Longer block time is not
favoured by block proposers, because it means less throughput of the system which usually means less overall rewards
for them. In the Argennon protocol the management of run-time resources, is left to the block proposer.

Resource usage can be measured per session or per application call. Obviously per session measurement is
easier and more efficient. However, when we measure resources per session if
a session violates its resource caps, determining the point of failure may require precise and error-free resource
measurement. For example, assume that a session containing an application call
violates a 2 milliseconds execution time cap. If in the caller's code, the call happens exactly after 2 milliseconds,
a small fluctuation in the execution time measurement can
change the point of failure between the called and the caller application. Note that for implementing optional
decoupling principle in Argennon, we need to determine the
exact application call which has failed in a call chain, and this fluctuation introduces a nondeterministic
behaviour which could make block validation impossible. That's why we need per application call
resource management sometimes. If we define the resource caps per application call and
perform our measurements for each application call separately, by using caps that are larger than the measurement
error, errors in the measurements would not change the point of failure in the call chain.

The AscEE has two type of execution sessions: \emph{optimistic} and \emph{monitored}. Resource usage of an optimistic
session is always measured per session and \textbf{default} pre-defined resource caps are used. On the other hand, for a
monitored session, resources that can not be measured precisely are measured per application call and their caps are
determined per application call by the external HTTP request (i.e.\ transaction). The block proposer decides the
type of the execution session for each transaction.

Different computational resources are measured and monitored during an AscEE session:
\begin{itemize}
    \item \textbf{execution time}:
    is the amount of cpu time that is required for executing a session or an application call. The execution time is
    measured in \emph{AscEE clocks}. One AscEE clock is defined as 1/1000 of the amount
    of \textbf{cpu time} needed for executing a predefined standard application which is used for benchmarking a host's
    performance. In other words, by definition, the AscEE imaginary standard machine completes the standard benchmark
    in 1000 clocks.

@@

In Argennon, execution time and local memory are considered \emph{nondeterministic} resources. A nondeterministic
resource is a resource that can not be measured precisely and its measurement always contains a random error.
Optimistic sessions are not allowed to fail because of violating nondeterministic resource limits. As a result, the
block proposer must always choose a monitored session for a transaction that is included in the block and violates a
nondeterministic resource cap.

When a transaction fails due to the violation of a limit for a nondeterministic resource, the proposer is
required to exactly specify the application call which violates that limit in the execution session. When validators are
executing a monitored session, for each application call, they enforce considerably larger limits for
nondeterministic resources. Only in case the proposer has declared that an application call violates a limit, the
validators will enforce the actual value of the limit. This simple mechanism ensures that, with a very high
probability, validators agree with the proposer, although their measurements could be different.

@@

In Argennon, some applications (smart contracts) are updatable. The ArgC Standard Library is an updatable smart
contract which can be updated by the Argennon governance
system. This means that bugs or security vulnerabilities in the ArgC Standard Library could be quickly patched and
applications could benefit from bugfixes and improvements of the ArgC Standard Library even if they are
non-updatable. Many important and useful functionalities,
such as fungible and non-fungible assets, access control mechanisms,
and general purpose DAOs are implemented in the ArgC Standard Library.
